---
title: "Visualization Methods in Data Analysis by R"
subtitle: "Laboratory Works Portfolio - Group 4, Variant 5"
author: "[Your Name]"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    embed-resources: true
execute:
  warning: false
  message: false
---

## Introduction

This report contains all laboratory works for the course
"Visualization Methods in Data Analysis by R" (2026).

```{r setup}
library(conflicted)
library(tidyverse)
library(patchwork)
library(MASS)
library(palmerpenguins)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Lab 0: R Setup and Basics

### Environment Setup

Installed R, RStudio, and the following key packages:
tidyverse, ggplot2, patchwork, GGally, naniar, plotly, ggiraph,
paletteer, colorBlindness, colorspace, lintr, styler.

### SpeedSki Dataset

```{r lab0-speedski}
library(GDAdata)

ggplot(SpeedSki, aes(Speed, fill = Sex)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.7) +
  facet_grid(rows = vars(Event)) +
  labs(title = "Speed Ski: Distribution by Event and Sex",
       x = "Speed (km/h)", y = "Count") +
  theme_minimal()
```

### Iris Dataset

```{r lab0-iris}
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
  geom_point(size = 2) +
  labs(title = "Iris: Petal Dimensions by Species") +
  theme_minimal()
```

### UCB Admissions

```{r lab0-ucb}
ucb_df <- as.data.frame(UCBAdmissions)

p1 <- ggplot(ucb_df, aes(Dept, weight = Freq)) +
  geom_bar(fill = "steelblue") + labs(title = "By Dept") + theme_minimal()
p2 <- ggplot(ucb_df, aes(Gender, weight = Freq)) +
  geom_bar(fill = "coral") + labs(title = "By Gender") + theme_minimal()
p3 <- ggplot(ucb_df, aes(Admit, weight = Freq)) +
  geom_bar(fill = "seagreen") + labs(title = "Admission") + theme_minimal()

p1 + p2 + p3 + plot_layout(widths = c(3, 1, 1))
```

## Lab 1: Exercises from "Graphical Data Analysis with R"

*Exercise 5 from each chapter 3-11.*

### Chapter 3: Zuni Educational Funding

```{r lab1-ch3}
library(lawstat)
data(zuni)

q05 <- quantile(zuni$revenue, 0.05)
q95 <- quantile(zuni$revenue, 0.95)

ggplot(zuni, aes(x = revenue)) +
  geom_histogram(binwidth = 500, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(q05, q95), color = "red", linetype = "dashed") +
  labs(title = "Zuni School Districts: Revenue per Pupil",
       subtitle = "Red lines = 5th and 95th percentiles",
       x = "Revenue per pupil ($)", y = "Count") +
  theme_minimal()
```

### Chapter 10: Diamonds Color vs Price

```{r lab1-ch10}
ggplot(diamonds, aes(x = color, y = price, fill = color)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(title = "Diamond Price by Color (D=best, J=worst)",
       subtitle = "Worse colors have higher prices due to confounding with carat",
       x = "Color", y = "Price ($)") +
  theme_minimal()
```

## Lab 2: autoGEDA

*Automated graphical exploratory data analysis using dlookr, brinton, plotluck, and manual approaches.*

```{r lab2}
data(penguins)
penguins_clean <- penguins |> drop_na()

numeric_vars <- penguins_clean |> select(where(is.numeric)) |> names()

plots <- map(numeric_vars, function(var) {
  ggplot(penguins_clean, aes(x = .data[[var]])) +
    geom_histogram(fill = "steelblue", color = "white", bins = 25) +
    labs(title = var) + theme_minimal()
})

wrap_plots(plots, ncol = 3)
```

## Lab 3: GEDA + Analytic Questions

### Elementary Questions

```{r lab3-elementary}
# IDENTIFY: Average body mass of Gentoo
ggplot(penguins_clean, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  labs(title = "IDENTIFY: Body mass by species",
       x = "Species", y = "Body Mass (g)") +
  theme_minimal()
```

```{r lab3-connect}
# CONNECT: Bill length vs bill depth
ggplot(penguins_clean, aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "CONNECT: Bill length vs depth (Simpson's paradox)",
       x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

### Synoptic Questions

```{r lab3-group}
# GROUP: Species clusters
ggplot(penguins_clean, aes(bill_length_mm, flipper_length_mm,
                           color = species, shape = species)) +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  labs(title = "GROUP: Three distinct species clusters",
       x = "Bill Length (mm)", y = "Flipper Length (mm)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

## Lab 4: Heat Map

```{r lab4}
# Correlation heatmap
cor_matrix <- airquality |>
  select(Ozone, Solar.R, Wind, Temp) |>
  drop_na() |>
  cor()

cor_long <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

ggplot(cor_long, aes(var1, var2, fill = correlation)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = round(correlation, 2)), size = 5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral",
                       midpoint = 0) +
  labs(title = "Correlation Heatmap: Air Quality") +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  coord_fixed()
```

## Lab 5: The Ugliest Plot

```{r lab5, fig.width=14, fig.height=10}
ugly_plot <- ggplot(penguins_clean, aes(bill_length_mm, body_mass_g)) +
  geom_line(aes(color = flipper_length_mm), linewidth = 2, alpha = 0.5) +
  geom_point(aes(size = bill_depth_mm), color = "yellow", stroke = 3) +
  scale_color_gradientn(colors = rainbow(20)) +
  scale_y_reverse(limits = c(6500, 2500)) +
  labs(title = "AMAZING PENGUIN ANALYSIS!!!",
       subtitle = "This proves penguins are DEFINITELY getting bigger*",
       caption = "*not actually tested") +
  theme(
    plot.background = element_rect(fill = "yellow", color = "red", linewidth = 5),
    panel.background = element_rect(fill = "lavender"),
    panel.grid.major = element_line(color = "red", linewidth = 1.5),
    plot.title = element_text(size = 24, color = "purple", face = "bold.italic",
                              hjust = 0.5)
  ) +
  annotate("text", x = 50, y = 4000, label = "WOW!", size = 20,
           color = "red", alpha = 0.3, angle = 30, fontface = "bold")

ugly_plot
```

## Lab 6: Visualization Recreation

```{r lab6}
pathogens <- read_csv("../data-I2Rsw.csv") |>
  rename(pathogen = Pathogen, illnesses = Illnesses,
         low = `Low Estimate`, high = `High estimate`) |>
  mutate(pathogen = fct_reorder(pathogen, illnesses))

ggplot(pathogens, aes(y = pathogen)) +
  geom_segment(aes(x = low, xend = high, yend = pathogen),
               color = "#D9D9D9", linewidth = 2) +
  geom_point(aes(x = low), color = "#E87722", size = 3) +
  geom_point(aes(x = illnesses), color = "#E87722", size = 5) +
  geom_point(aes(x = high), color = "#E87722", size = 3) +
  scale_x_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  labs(title = "Foodborne Illness Estimates by Pathogen (CDC 2019)",
       subtitle = "Ranges show 90% credible intervals",
       x = "", y = "") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

## Labs 7-9: Munzner Foundations & Color

*These labs are completed as separate PDF worksheets using the Google Drive templates.*

- **Lab 7**: Foundations 1 - Data Abstraction, Palmer Penguins, Foreign Aid
- **Lab 8**: Foundations 2 - Marks & Channels, Generate Two Numbers, Redesign
- **Lab 9**: Complexity Families + Color palettes and accessibility

## Session Info

```{r}
sessionInfo()
```
