---
title: "Lab 0: Install R, RStudio + Packages"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    embed-resources: true
execute:
  warning: false
  message: false
---

## Setup

```{r setup}
library(conflicted)
library(tidyverse)
library(patchwork)
library(MASS)

conflicts_prefer(dplyr::filter, dplyr::select, dplyr::lag)
```

## 1. Tidyverse Style Guide

Following the [tidyverse style guide](https://style.tidyverse.org/):

- `snake_case` for names
- `<-` for assignment
- Spaces around operators
- Max 80 characters per line
- `|>` pipe for chaining
- `styler::style_file()` to auto-format
- `lintr::lint()` to check

## 2. SpeedSki Dataset

```{r speedski}
library(GDAdata)

glimpse(SpeedSki)
summary(SpeedSki)
```

### Histogram of Speed

```{r speedski-hist}
ggplot(SpeedSki, aes(Speed)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Speed Ski Record Speeds",
    x = "Speed (km/h)", y = "Count"
  ) +
  theme_minimal()
```

### Faceted by Sex and Event

```{r speedski-facet}
ggplot(SpeedSki, aes(Speed, fill = Sex)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.7) +
  facet_grid(rows = vars(Event)) +
  labs(title = "Speed Distribution by Event and Sex",
       x = "Speed (km/h)", y = "Count") +
  theme_minimal()
```

### Automated Plotting with purrr + patchwork

```{r speedski-purrr}
library(purrr)

names1 <- levels(SpeedSki$Event)

plot1 <- function(data, var1, var2, var3) {
  data |>
    filter({{ var2 }} %in% var3) |>
    ggplot(aes(x = {{ var1 }})) +
    geom_histogram(fill = "steelblue", color = "white", binwidth = 5) +
    labs(title = var3) +
    theme_minimal()
}

group1 <- map(names1, ~plot1(SpeedSki, Speed, Event, .x))

wrap_plots(group1, ncol = 1) &
  lims(x = c(160, 215), y = c(0, 9))
```

## 3. Iris Dataset

```{r iris}
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
  geom_point(size = 2) +
  labs(title = "Iris: Petal Length vs Petal Width",
       x = "Petal Length (cm)", y = "Petal Width (cm)") +
  theme_minimal()
```

## 4. UCB Admissions

```{r ucb}
ucb_df <- as.data.frame(UCBAdmissions)

p1 <- ggplot(ucb_df, aes(Dept, weight = Freq)) +
  geom_bar(fill = "steelblue") +
  labs(title = "By Department") + theme_minimal()

p2 <- ggplot(ucb_df, aes(Gender, weight = Freq)) +
  geom_bar(fill = "coral") +
  labs(title = "By Gender") + theme_minimal()

p3 <- ggplot(ucb_df, aes(Admit, weight = Freq)) +
  geom_bar(fill = "seagreen") +
  labs(title = "Admission Status") + theme_minimal()

p1 + p2 + p3 + plot_layout(widths = c(3, 1, 1))
```

## 5. Pima Indians Diabetes

```{r pima-hist}
pima_vars <- c("glu", "bp", "skin", "bmi", "ped", "age")

plots_hist <- map(pima_vars, function(var) {
  ggplot(Pima.tr2, aes(x = .data[[var]])) +
    geom_histogram(fill = "steelblue", color = "white", bins = 20) +
    labs(title = var) + theme_minimal()
})

wrap_plots(plots_hist, ncol = 3)
```

```{r pima-box}
plots_box <- map(pima_vars, function(var) {
  ggplot(Pima.tr2, aes(y = .data[[var]])) +
    geom_boxplot(fill = "steelblue", alpha = 0.5) +
    labs(title = var) + theme_minimal()
})

wrap_plots(plots_box, nrow = 1)
```

## 6. Titanic

```{r titanic}
titanic_df <- as.data.frame(Titanic)

tp1 <- ggplot(titanic_df, aes(Class, weight = Freq)) +
  geom_bar(fill = "steelblue") + theme_minimal()
tp2 <- ggplot(titanic_df, aes(Sex, weight = Freq)) +
  geom_bar(fill = "coral") + theme_minimal()
tp3 <- ggplot(titanic_df, aes(Age, weight = Freq)) +
  geom_bar(fill = "gold3") + theme_minimal()
tp4 <- ggplot(titanic_df, aes(Survived, weight = Freq)) +
  geom_bar(fill = "seagreen") + theme_minimal()

(tp1 + tp2 + tp3 + tp4) +
  plot_layout(nrow = 1, widths = c(2, 1, 1, 1)) &
  scale_y_continuous(limits = c(0, 2200))
```

## 7. Old Faithful Geyser

```{r faithful}
ggplot(faithful, aes(eruptions, waiting)) +
  geom_point(alpha = 0.5) +
  geom_density_2d(color = "steelblue") +
  labs(title = "Old Faithful: Eruption Duration vs Waiting Time",
       x = "Eruption duration (min)", y = "Waiting time (min)") +
  theme_minimal()
```

## 8. London Bikes

```{r bikes}
library(readr)
library(lubridate)

bikes <- read_csv("../london-bikes.csv", col_types = "Dcfffilllddddc",
                   show_col_types = FALSE)

temp_locale <- Sys.getlocale("LC_TIME")
Sys.setlocale("LC_TIME", "C")

ggplot(bikes, aes(x = month(date, label = TRUE), y = temp)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  labs(title = "London Bike Sharing: Temperature by Month",
       x = "Month", y = "Temperature (°C)") +
  theme_minimal()

Sys.setlocale("LC_TIME", temp_locale)
```

## Session Info

```{r}
sessionInfo()
```
