---
title: "Lab 9: Complexity Families + Color"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(patchwork)
library(colorspace)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Part 1: Complexity Families (Munzner)

### Dataset types

| Family | Examples | Key Structure |
|:-------|:---------|:--------------|
| **Tables** | Flat table, multidimensional table | Rows = items, columns = attributes |
| **Networks** | Trees, general graphs | Nodes + links |
| **Spatial** | Fields (continuous), geometry | Spatial position is data |
| **Sets** | Lists, clusters, partitions | Membership relationships |

### Classifying datasets

```{r complexity-example}
library(palmerpenguins)
data(penguins)
penguins_clean <- penguins |> drop_na()

# TABLE example: Palmer Penguins
ggplot(penguins_clean, aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point(alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Table Data: Palmer Penguins",
       subtitle = "Flat table with quantitative and categorical attributes",
       x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  theme_minimal()
```

```{r network-example}
# NETWORK example: simple hierarchy (simulated)
tree_df <- tibble(
  from = c("Root", "Root", "A", "A", "B", "B", "C", "C"),
  to = c("A", "B", "C", "D", "E", "F", "G", "H"),
  level = c(1, 1, 2, 2, 2, 2, 3, 3)
)

ggplot(tree_df, aes(x = from, y = to)) +
  geom_point(size = 3, color = "steelblue") +
  geom_segment(aes(xend = to, yend = from), color = "gray60") +
  labs(title = "Network Data: Tree Structure",
       subtitle = "Nodes connected by parent-child links") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Part 2: Color Spaces

### RGB vs HCL

- **RGB**: Device-oriented, not perceptually uniform
- **HCL**: Hue-Chroma-Luminance, perceptually uniform (preferred)

```{r color-spaces, fig.height=4}
# Compare color spaces
hcl_pal <- hcl.colors(9, "viridis")
rainbow_pal <- rainbow(9)

par(mfrow = c(2, 1), mar = c(2, 4, 2, 1))
barplot(rep(1, 9), col = hcl_pal, border = NA, main = "HCL: Viridis (perceptually uniform)")
barplot(rep(1, 9), col = rainbow_pal, border = NA, main = "RGB: Rainbow (NOT uniform)")
```

### Spectral properties

```{r specplot, fig.height=8}
specplot(hcl.colors(9, "viridis"), main = "Viridis")
```

## Part 3: Color Palette Types

### Sequential palettes

For ordered data (low to high):

```{r sequential}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis") +
  labs(title = "Sequential: Viridis",
       subtitle = "One hue, varying luminance") +
  theme_minimal()
```

### Diverging palettes

For data with meaningful midpoint:

```{r diverging}
cor_matrix <- penguins_clean |>
  select(where(is.numeric)) |>
  cor()

cor_long <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r")

ggplot(cor_long, aes(var1, var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral",
                       midpoint = 0) +
  labs(title = "Diverging: Blue-White-Red",
       subtitle = "Midpoint at 0 for correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed()
```

### Qualitative palettes

For nominal categories:

```{r qualitative}
ggplot(penguins_clean, aes(species, fill = species)) +
  geom_bar(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Qualitative: Set2",
       subtitle = "Different hues, similar luminance for categories") +
  theme_minimal()
```

## Part 4: Color Accessibility

- ~8% of males and ~0.5% of females have color vision deficiency
- Always test with colorblind simulators
- Avoid red-green as sole distinguishing colors
- Use redundant encodings (shape + color)

```{r accessibility}
# Good practice: redundant encoding
ggplot(penguins_clean, aes(bill_length_mm, flipper_length_mm,
                           color = species, shape = species)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Accessible: Color + Shape Redundancy",
       subtitle = "Distinguishable even without color",
       x = "Bill Length (mm)", y = "Flipper Length (mm)") +
  theme_minimal()
```

### Simulating color blindness

```{r cvd-simulation, fig.height=8}
base_colors <- RColorBrewer::brewer.pal(3, "Set2")

p <- ggplot(penguins_clean, aes(bill_length_mm, flipper_length_mm, color = species)) +
  geom_point(alpha = 0.7) +
  labs(x = "Bill Length (mm)", y = "Flipper Length (mm)") +
  theme_minimal()

p_normal <- p + scale_color_manual(values = base_colors) +
  labs(title = "Normal vision")
p_deutan <- p + scale_color_manual(values = deutan(base_colors)) +
  labs(title = "Deuteranopia simulation")
p_protan <- p + scale_color_manual(values = protan(base_colors)) +
  labs(title = "Protanopia simulation")
p_tritan <- p + scale_color_manual(values = tritan(base_colors)) +
  labs(title = "Tritanopia simulation")

(p_normal + p_deutan) / (p_protan + p_tritan)
```

## HCL Palette Browser

```{r hcl-palettes, fig.height=10}
hcl_palettes(plot = TRUE, n = 7)
```

## Conclusions

- Use **HCL-based palettes** (viridis, colorspace) for perceptual uniformity.
- Match palette type to data type: sequential, diverging, or qualitative.
- Always check accessibility with CVD simulations.
- Add redundant encodings (shape, text labels) for critical distinctions.

```{r session}
sessionInfo()
```
