---
title: "Lab 8: Tamara Munzner -- Foundations 2"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(patchwork)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Part 1: Generate -- Two Numbers

Given two numbers, we explore different visualization encodings ranked by effectiveness.

### Channel Effectiveness (Munzner)

1. Position on common scale (most effective)
2. Position on unaligned scale
3. Length
4. Tilt/Angle
5. Area
6. Color luminance
7. Color saturation

```{r two-numbers}
two_nums <- tibble(name = c("A", "B"), value = c(73, 42))

# Most effective: position on common scale (bar chart)
p1 <- ggplot(two_nums, aes(name, value, fill = name)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(title = "1. Position (best)", y = "Value", x = "") +
  theme_minimal()

# Length encoding
p2 <- ggplot(two_nums, aes(x = value, y = name, fill = name)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(title = "2. Length", x = "Value", y = "") +
  theme_minimal()

# Area encoding (bubble)
p3 <- ggplot(two_nums, aes(x = name, y = 1, size = value, color = name)) +
  geom_point(show.legend = FALSE) +
  scale_size_area(max_size = 25) +
  scale_color_manual(values = c("steelblue", "coral")) +
  labs(title = "3. Area (worse)") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

# Color luminance encoding
p4 <- ggplot(two_nums, aes(x = name, y = 1, fill = value)) +
  geom_tile(width = 0.8, height = 0.8, show.legend = TRUE) +
  geom_text(aes(label = value), size = 8, color = "white") +
  scale_fill_gradient(low = "gray90", high = "steelblue") +
  labs(title = "4. Color luminance (worst)") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "Encoding Two Numbers: Effectiveness Comparison")
```

Position on a common scale makes it easiest to compare the two values. Area and color luminance make precise comparison much harder.

## Part 2: Marks and Channels Analysis

### Marks (geometric primitives)

- **Points**: 0D marks for individual items
- **Lines**: 1D marks for connections or trends
- **Areas**: 2D marks for regions or categories

### Channels (visual variables)

- **Position**: Most effective for quantitative data
- **Color hue**: Best for categorical data (limited to ~12 categories)
- **Size**: Moderate effectiveness for quantitative data
- **Shape**: Categorical only, limited to ~7 distinct shapes

```{r marks-channels, fig.height=8}
library(palmerpenguins)
data(penguins)
penguins_clean <- penguins |> drop_na()

# Demonstrate marks and channels
p_points <- ggplot(penguins_clean, aes(bill_length_mm, bill_depth_mm)) +
  geom_point(color = "steelblue", alpha = 0.5) +
  labs(title = "Marks: Points", subtitle = "Channel: Position (x, y)") +
  theme_minimal()

p_lines <- ggplot(penguins_clean, aes(bill_length_mm, body_mass_g)) +
  geom_smooth(aes(color = species), method = "lm", se = FALSE) +
  labs(title = "Marks: Lines", subtitle = "Channels: Position + Color hue") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()

p_areas <- ggplot(penguins_clean, aes(species, fill = species)) +
  geom_bar(show.legend = FALSE, alpha = 0.8) +
  labs(title = "Marks: Areas (bars)", subtitle = "Channels: Position + Length") +
  theme_minimal()

p_combined <- ggplot(penguins_clean,
  aes(bill_length_mm, flipper_length_mm,
      color = species, shape = sex, size = body_mass_g)) +
  geom_point(alpha = 0.6) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Combined: Multiple channels",
       subtitle = "Position + Color + Shape + Size") +
  theme_minimal()

(p_points + p_lines) / (p_areas + p_combined)
```

## Part 3: Redesign -- New Zealand Table

### Principles for table redesign:

1. **Replace text with visual encodings** where appropriate
2. **Use position** (bars) for quantitative comparison
3. **Keep text** for categorical labels
4. **Add color coding** for quick scanning
5. **Sort by key metric** for faster lookup

```{r table-redesign}
library(gt)

# Example: NZ population data (simulated)
nz_data <- tibble(
  Region = c("Auckland", "Wellington", "Canterbury", "Waikato",
             "Bay of Plenty", "Manawatu", "Otago"),
  Population = c(1717500, 527500, 645900, 496700, 338300, 252300, 245600),
  Growth = c(2.1, 1.5, 1.2, 1.8, 2.3, 0.8, 1.1),
  MedianAge = c(34.2, 35.8, 37.1, 36.5, 38.9, 33.2, 36.7)
) |>
  arrange(desc(Population))

nz_data |>
  gt() |>
  tab_header(
    title = "New Zealand Regions: Key Demographics",
    subtitle = "Redesigned from tabular format with visual cues"
  ) |>
  fmt_number(columns = Population, suffixing = TRUE) |>
  fmt_percent(columns = Growth, scale_values = FALSE, decimals = 1) |>
  data_color(columns = Population, palette = c("white", "steelblue")) |>
  data_color(columns = Growth, palette = c("white", "seagreen")) |>
  cols_align(align = "center")
```

## Expressiveness and Effectiveness Principles

- **Expressiveness**: Encode all and only the information in the data.
- **Effectiveness**: Use the most effective channels for the most important attributes.

A bar chart is *expressive* (shows the quantitative values) and *effective* (uses position on a common scale, the most effective channel).

## Conclusions

- Position is the most effective channel for quantitative data.
- Multiple channels can be combined, but avoid overloading (max 3--4 simultaneous channels).
- Table redesign should replace text-heavy formats with visual encodings where possible.

```{r session}
sessionInfo()
```
