---
title: "Lab 1: Exercises from Graphical Data Analysis with R"
subtitle: "One exercise (Nr. 5) from each chapter 3-11"
author: "Group 4, Variant 5"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 6
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(conflicted)
library(tidyverse)
library(patchwork)
library(GGally)
library(MASS)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Chapter 3: Zuni Educational Funding

**Dataset:** `zuni` from `lawstat` — 89 school districts in New Mexico with revenue per pupil.

**Task:** Assess whether the lowest/highest 5% of revenue values are extreme; check symmetry and normality of trimmed distribution.

```{r ch3-overview}
library(lawstat)
data(zuni)

q05 <- quantile(zuni$revenue, 0.05)
q95 <- quantile(zuni$revenue, 0.95)

p_hist <- ggplot(zuni, aes(x = revenue)) +
  geom_histogram(binwidth = 500, fill = "steelblue", color = "white") +
  geom_vline(xintercept = c(q05, q95), color = "red", linetype = "dashed") +
  labs(title = "Revenue per Pupil", subtitle = "Red lines: 5th/95th percentiles",
       x = "Revenue ($)", y = "Count") +
  theme_minimal()

p_box <- ggplot(zuni, aes(y = revenue)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  labs(title = "Boxplot", y = "Revenue ($)") +
  theme_minimal()

p_hist + p_box + plot_layout(widths = c(3, 1))
```

**(a)** The histogram is preferred — it shows the right-skewed shape with a few very high values. The boxplot marks these as outliers but hides the distribution form.

```{r ch3-density}
zuni_trimmed <- zuni |> filter(revenue >= q05, revenue <= q95)

ggplot(zuni_trimmed, aes(x = revenue)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 300,
                 fill = "steelblue", color = "white", alpha = 0.5) +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Trimmed Distribution (5%-95%)",
       subtitle = "Roughly symmetric after removing extremes",
       x = "Revenue ($)", y = "Density") +
  theme_minimal()
```

**(b)** After trimming, the distribution looks roughly symmetric.

```{r ch3-qq}
ggplot(zuni_trimmed, aes(sample = revenue)) +
  stat_qq(color = "steelblue") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Trimmed Revenue",
       subtitle = "Close to normal with minor tail deviations",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```

**(c)** The Q-Q plot shows approximate normality with slight deviations in the tails.

## Chapter 4: Occupational Mobility

**Dataset:** `Yamaguchi87` from `vcdExtra` — occupational mobility tables for UK, US, Japan.

```{r ch4}
library(vcdExtra)
data(Yamaguchi87)
yam_df <- as.data.frame(Yamaguchi87)

sons_by_country <- yam_df |>
  group_by(Country, Son) |>
  summarise(n = sum(Freq), .groups = "drop") |>
  group_by(Country) |>
  mutate(prop = n / sum(n))

ggplot(sons_by_country, aes(x = Son, y = prop, fill = Country)) +
  geom_col(position = "dodge") +
  labs(title = "Sons' Occupations by Country",
       x = "Occupation", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**(a)** Japan has a distinctive pattern — more farm workers, fewer upper non-manual than UK/US.

```{r ch4-uk}
uk_data <- yam_df |> filter(Country == "UK")

uk_gen <- bind_rows(
  uk_data |> group_by(occ = Father) |> summarise(n = sum(Freq), .groups = "drop") |>
    mutate(gen = "Father"),
  uk_data |> group_by(occ = Son) |> summarise(n = sum(Freq), .groups = "drop") |>
    mutate(gen = "Son")
) |> group_by(gen) |> mutate(prop = n / sum(n))

ggplot(uk_gen, aes(x = occ, y = prop, fill = gen)) +
  geom_col(position = "dodge") +
  labs(title = "UK: Fathers' vs Sons' Occupations",
       subtitle = "Shift from manual to non-manual work across generations",
       x = "Occupation", y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**(b-c)** Clear upward mobility — sons move toward non-manual occupations. Expected given post-war economic development.

## Chapter 5: Bank Sex Discrimination

**Dataset:** `case1202` from `Sleuth2` — employment variables (seniority, age, experience).

```{r ch5}
library(Sleuth2)
data(case1202)

case1202 |>
  select(Senior, Age, Exper) |>
  ggpairs(
    lower = list(continuous = wrap("points", alpha = 0.4, color = "steelblue")),
    diag = list(continuous = wrap("densityDiag", fill = "steelblue", alpha = 0.4)),
    upper = list(continuous = wrap("cor", size = 5))
  ) +
  labs(title = "Scatterplot Matrix: Bank Employment Variables") +
  theme_minimal()
```

**(a)** Age and Experience are strongly correlated (both accumulate with time). Senior shows a triangular pattern with Age — you can't have more seniority than years since hiring.

**(b)** Seniority depends on *when* someone was hired, not their total life span. That's why Senior vs Age doesn't show the same linear structure as Experience vs Age.

## Chapter 6: Bodyfat

**Dataset:** `bodyfat` from `mfp` — 252 men with body measurements.

```{r ch6}
library(mfp)
data(bodyfat)

ggparcoord(bodyfat, columns = 1:ncol(bodyfat), alpha = 0.08,
           scale = "uniminmax") +
  labs(title = "Parallel Coordinate Plot: Bodyfat Data",
       subtitle = "Scaled to [0,1]. Look for outliers and variable relationships.",
       x = "Variable", y = "Scaled Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
```

**(a)** Several outliers visible — lines that cross extreme values on multiple axes.

**(b)** One case with extremely low height stands out — likely a data entry error (height recorded in cm instead of inches?).

**(c)** Density and bodyfat show a strong inverse relationship — visible as crossing lines between these two adjacent axes.

**(d)** Grouping related circumference measurements together would reveal body type patterns more clearly.

## Chapter 7: Airline Arrivals (Simpson's Paradox)

**Dataset:** Alaska Airlines vs America West at 5 airports.

```{r ch7}
airline_data <- tibble(
  airline = rep(c("Alaska", "America West"), each = 5),
  airport = rep(c("Los Angeles", "Phoenix", "San Diego",
                  "San Francisco", "Seattle"), 2),
  on_time = c(497, 221, 212, 503, 1841, 694, 4840, 383, 320, 201),
  delayed = c(62, 12, 20, 102, 305, 117, 415, 65, 129, 61)
) |> mutate(total = on_time + delayed, delay_rate = delayed / total)

overall <- airline_data |>
  group_by(airline) |>
  summarise(rate = sum(delayed) / sum(total), .groups = "drop")

p_overall <- ggplot(overall, aes(airline, rate, fill = airline)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(title = "Overall Delay Rate", x = "", y = "Delay Rate") +
  theme_minimal()

p_by_airport <- ggplot(airline_data, aes(airline, delay_rate, fill = airline)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~airport, nrow = 1) +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(title = "Delay Rate by Airport",
       subtitle = "Alaska is better at EVERY airport! (Simpson's Paradox)",
       x = "", y = "Delay Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_overall / p_by_airport
```

Alaska flies mostly to Seattle (high delay), America West mostly to Phoenix (low delay). Aggregation reverses the per-airport advantage.

## Chapter 8: Pima Indians Overview

**Dataset:** `Pima.tr` + `Pima.te` from `MASS`.

```{r ch8}
pima_r <- bind_rows(Pima.tr, Pima.te)

pima_long <- pima_r |>
  pivot_longer(c(glu, bp, skin, bmi, ped, age),
               names_to = "variable", values_to = "value")

ggplot(pima_long, aes(x = value, fill = type)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  facet_wrap(~variable, scales = "free") +
  scale_fill_manual(values = c("steelblue", "coral")) +
  labs(title = "Pima Indians: Continuous Variables by Diabetes Type",
       x = "Value", y = "Count") +
  theme_minimal()
```

Diabetic individuals (Yes) tend to have higher glucose, BMI, age, and number of pregnancies.

## Chapter 9: Pearson Heights — Outliers

**Dataset:** `father.son` from `UsingR`.

```{r ch9}
library(UsingR)
data(father.son)

p_scatter <- ggplot(father.son, aes(fheight, sheight)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Father-Son Heights (Pearson)",
       subtitle = "Red: linear fit, Dashed: y = x",
       x = "Father's Height (in)", y = "Son's Height (in)") +
  theme_minimal()

p_density <- ggplot(father.son, aes(fheight, sheight)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_density_2d(color = "red") +
  labs(title = "With Bivariate Density Contours",
       subtitle = "Points outside outermost contour = potential outliers",
       x = "Father's Height (in)", y = "Son's Height (in)") +
  theme_minimal()

p_scatter + p_density
```

**(a)** Cases in the corners (short father + tall son, or vice versa) are potential outliers.

**(b)** The density contours identify a slightly different set — bivariate outliers that are far from the main concentration even if not extreme on individual axes.

## Chapter 10: Diamonds — Color vs Price

**Dataset:** `diamonds` from `ggplot2`.

```{r ch10-boxplot}
ggplot(diamonds, aes(x = color, y = price, fill = color)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_fill_viridis_d() +
  labs(title = "Diamond Price by Color (D=best, J=worst)",
       subtitle = "Surprise: 'worse' colors have higher median prices!",
       x = "Color", y = "Price ($)") +
  theme_minimal()
```

**(a)** This is confounded by carat — worse-colored diamonds tend to be larger. Larger diamonds cost more regardless of color.

```{r ch10-coef}
model <- lm(log(price) ~ color + log(carat), data = diamonds)
coef_df <- broom::tidy(model, conf.int = TRUE) |>
  filter(str_starts(term, "color"))

ggplot(coef_df, aes(x = term, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_pointrange(color = "steelblue", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Color Coefficients (controlling for carat)",
       subtitle = "After controlling for size, worse colors DO reduce price",
       x = "", y = "Coefficient (log price)") +
  theme_minimal()
```

**(b)** After controlling for carat, the expected pattern emerges: each step from D toward J decreases price.

## Chapter 11: Bundesliga Goals

**Dataset:** `Bundesliga` from `vcd`.

```{r ch11-trellis}
library(vcd)
data(Bundesliga)

bl_summary <- Bundesliga |>
  mutate(season = ifelse(as.numeric(format(Date, "%m")) >= 7,
                         as.numeric(format(Date, "%Y")),
                         as.numeric(format(Date, "%Y")) - 1)) |>
  group_by(season, HomeTeam) |>
  summarise(avg_goals = mean(HomeGoals), games = n(), .groups = "drop")

top_teams <- bl_summary |> count(HomeTeam, sort = TRUE) |>
  filter(n >= 30) |> pull(HomeTeam)

ggplot(bl_summary |> filter(HomeTeam %in% top_teams),
       aes(season, avg_goals, color = HomeTeam)) +
  geom_line(alpha = 0.7) +
  geom_smooth(se = FALSE, linewidth = 0.5) +
  facet_wrap(~HomeTeam) +
  labs(title = "Average Home Goals per Game (Bundesliga)",
       subtitle = "Teams with 30+ seasons. Trellis display handles gaps naturally.",
       x = "Season", y = "Avg Home Goals") +
  theme_minimal() +
  theme(legend.position = "none")
```

**(a)** Trellis display is best — it handles missing years for teams that were relegated. Hamburg is the only ever-present team.

```{r ch11-comparison}
home_g <- Bundesliga |>
  mutate(season = ifelse(as.numeric(format(Date, "%m")) >= 7,
                         as.numeric(format(Date, "%Y")),
                         as.numeric(format(Date, "%Y")) - 1)) |>
  filter(HomeTeam %in% c("Hamburger SV", "Bayern Munich")) |>
  group_by(season, team = HomeTeam) |>
  summarise(avg = mean(HomeGoals), .groups = "drop") |>
  mutate(loc = "Home")

away_g <- Bundesliga |>
  mutate(season = ifelse(as.numeric(format(Date, "%m")) >= 7,
                         as.numeric(format(Date, "%Y")),
                         as.numeric(format(Date, "%Y")) - 1)) |>
  filter(AwayTeam %in% c("Hamburger SV", "Bayern Munich")) |>
  group_by(season, team = AwayTeam) |>
  summarise(avg = mean(AwayGoals), .groups = "drop") |>
  mutate(loc = "Away")

hb <- bind_rows(home_g, away_g)

ggplot(hb, aes(season, avg, color = loc)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~team) +
  scale_color_manual(values = c("steelblue", "coral")) +
  labs(title = "Home vs Away Scoring: Hamburg & Bayern",
       x = "Season", y = "Avg Goals/Game", color = "") +
  theme_minimal()
```

**(b)** Time series shows trends; a scatterplot would show correlation between home/away performance. Both are informative in different ways.

## Session Info

```{r}
sessionInfo()
```
