---
title: "Lab 4: Heat Map + Table Visualization"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(patchwork)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Dataset

We use the built-in `airquality` dataset (New York, May--September 1973) with Ozone, Solar Radiation, Wind, and Temperature measurements. This is supplemented with a TidyTuesday-style exploration.

```{r data-prep}
data(airquality)

aq_matrix <- airquality |>
  select(Ozone, Solar.R, Wind, Temp) |>
  drop_na() |>
  as.matrix() |>
  scale()

glimpse(airquality)
```

## Heatmap 1: ggplot2 geom\_tile()

The most flexible approach --- full ggplot2 grammar for customization.

```{r heatmap-tile, fig.height=5}
ggplot(airquality, aes(x = Day, y = factor(Month, labels = month.abb[5:9]),
                       fill = Ozone)) +
  geom_tile(color = "white", size = 0.3) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90",
                       name = "Ozone\n(ppb)") +
  labs(title = "Daily Ozone Levels (New York, 1973)",
       subtitle = "Built with ggplot2::geom_tile()",
       x = "Day of Month", y = "Month") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Ozone peaks in July--August. Missing values (grey) are concentrated in June.

## Heatmap 2: Base R heatmap()

Includes dendrograms by default, showing hierarchical clustering.

```{r heatmap-base, fig.height=6}
heatmap(
  aq_matrix[1:50, ],
  col = hcl.colors(50, "YlOrRd", rev = TRUE),
  scale = "none",
  main = "Base R heatmap() - Air Quality (first 50 obs)"
)
```

## Heatmap 3: pheatmap

Better aesthetics, annotation support, and clustering control than base R.

```{r heatmap-pheatmap, fig.height=7}
library(pheatmap)

pheatmap(
  aq_matrix[1:50, ],
  color = hcl.colors(50, "Viridis"),
  main = "pheatmap - Air Quality Data (scaled)",
  fontsize = 10,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE
)
```

## Heatmap 4: gplots::heatmap.2()

Extended version of base `heatmap()` with color key, density info, and trace lines.

```{r heatmap-gplots, fig.height=7}
library(gplots)

heatmap.2(
  aq_matrix[1:50, ],
  col = hcl.colors(50, "Viridis"),
  scale = "none",
  trace = "none",
  dendrogram = "both",
  margins = c(8, 4),
  main = "gplots::heatmap.2() - Air Quality",
  key.title = "Scaled Value",
  labRow = FALSE
)
```

## Heatmap 5: Correlation Heatmap

```{r heatmap-correlation, fig.height=5, fig.width=7}
cor_matrix <- airquality |>
  select(Ozone, Solar.R, Wind, Temp) |>
  drop_na() |>
  cor()

cor_long <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = round(correlation, 2)), color = "black", size = 5) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Heatmap - Air Quality Variables",
       x = "", y = "") +
  theme_minimal() +
  theme(panel.grid = element_blank(), axis.text = element_text(size = 12)) +
  coord_fixed()
```

Ozone and Temperature: strong positive (r = 0.70). Wind: negatively correlated with both.

## Heatmap 6: Calendar Heatmap

```{r heatmap-calendar, fig.height=4}
aq_cal <- airquality |>
  mutate(
    Date = as.Date(paste(1973, Month, Day, sep = "-")),
    week = as.numeric(format(Date, "%U")),
    wday = factor(weekdays(Date, abbreviate = TRUE),
                  levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))
  )

ggplot(aq_cal, aes(x = week, y = fct_rev(wday), fill = Temp)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_viridis_c(option = "inferno", name = "Temp (\u00B0F)") +
  labs(title = "Calendar Heatmap: Temperature (May-Sep 1973)",
       x = "Week of Year", y = "") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## Color Interpolation and Class Intervals

Different methods for binning continuous data into classes
(Lambert & Zanin, *Practical Handbook of Thematic Cartography*, p.49):

- **Linear** (equal intervals): simple, but sensitive to outliers
- **Quantiles**: each bin has equal count of observations
- **Natural breaks** (Jenks): minimizes within-class variance
- **Standard deviation**: bins centered on the mean

Reference: [How to choose an interpolation for your color scale](https://blog.datawrapper.de/interpolation-for-color-scales-and-maps/)

```{r class-intervals, fig.height=5}
ozone_clean <- na.omit(airquality$Ozone)

# Compare binning methods
breaks_equal <- seq(min(ozone_clean), max(ozone_clean), length.out = 7)
breaks_quantile <- quantile(ozone_clean, probs = seq(0, 1, length.out = 7))

p_equal <- ggplot(data.frame(Ozone = ozone_clean), aes(x = Ozone)) +
  geom_histogram(breaks = breaks_equal, fill = "steelblue", color = "white") +
  geom_vline(xintercept = breaks_equal, color = "red", linetype = "dashed", size = 0.5) +
  labs(title = "Equal Intervals", y = "Count") +
  theme_minimal()

p_quantile <- ggplot(data.frame(Ozone = ozone_clean), aes(x = Ozone)) +
  geom_histogram(breaks = as.numeric(breaks_quantile),
                 fill = "coral", color = "white") +
  geom_vline(xintercept = as.numeric(breaks_quantile),
             color = "red", linetype = "dashed", size = 0.5) +
  labs(title = "Quantile Intervals", y = "Count") +
  theme_minimal()

p_equal + p_quantile +
  plot_annotation(title = "Class Interval Methods for Ozone Data")
```

## Table Visualization

### knitr::kable()

The simplest table output for R Markdown / PDF.

```{r table-kable}
aq_summary <- airquality |>
  group_by(Month = month.abb[Month]) |>
  summarise(
    `Mean Ozone` = round(mean(Ozone, na.rm = TRUE), 1),
    `Mean Temp` = round(mean(Temp), 1),
    `Mean Wind` = round(mean(Wind), 1),
    `Mean Solar` = round(mean(Solar.R, na.rm = TRUE), 1),
    Days = n(),
    .groups = "drop"
  )

knitr::kable(aq_summary, caption = "Air Quality Summary (1973) -- knitr::kable()")
```

### kableExtra

Enhanced tables with styling, striped rows, and conditional formatting.

```{r table-kableextra}
library(kableExtra)

aq_summary |>
  kbl(caption = "Air Quality Summary -- kableExtra") |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center",
    latex_options = c("striped", "hold_position")
  ) |>
  column_spec(2, bold = TRUE,
              color = ifelse(aq_summary$`Mean Ozone` > 50, "red", "black")) |>
  row_spec(0, bold = TRUE, color = "white", background = "#4682B4")
```

### formattable (concept)

```{r table-formattable}
library(formattable)

formattable(aq_summary, list(
  `Mean Ozone` = color_bar("coral"),
  `Mean Temp` = color_bar("steelblue"),
  `Mean Wind` = color_tile("white", "seagreen")
))
```

## Table Packages Overview

| Package | Static PDF | Interactive HTML | Best For |
|:--------|:----------:|:---------------:|:---------|
| `knitr::kable` | Yes | No | Simple tables |
| `kableExtra` | Yes | Yes | Styled PDF/HTML tables |
| `gt` | Yes | Yes | Publication-quality tables |
| `gtExtras` | No | Yes | Sparklines, plots in tables |
| `DT` | No | Yes | Interactive data exploration |
| `reactable` | No | Yes | Sorting, filtering, pagination |
| `formattable` | Partial | Yes | Conditional formatting |
| `flextable` | Yes | Yes | Word/PowerPoint output |

## Color Palette Notes

- **Sequential** palettes (viridis, plasma, inferno) for ordered continuous data
- **Diverging** palettes (RdBu, PiYG) for data with meaningful midpoint
- **Qualitative** palettes (Set2, Dark2) for nominal categories
- Always use **colorblind-friendly** palettes (viridis family)

```{r session}
sessionInfo()
```
