---
title: "Lab 5: The Ugliest Plot + Visualization Misleaders"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(patchwork)
library(palmerpenguins)

conflicts_prefer(dplyr::filter, dplyr::select)

data(penguins)
penguins_clean <- penguins |> drop_na()
```

## The Ugliest Plot Possible

This plot intentionally violates as many visualization principles as possible, based on CALVI's 11 misleaders: inverted axis, rainbow colors, inappropriate geom, cluttered annotations, distorted aspect ratio, and terrible theme choices.

```{r ugly-plot, fig.width=14, fig.height=10}
ugly_plot <- ggplot(penguins_clean, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_line(aes(color = flipper_length_mm), size = 2, alpha = 0.5) +
  geom_point(aes(size = bill_depth_mm), color = "yellow", stroke = 3) +
  scale_color_gradientn(colors = rainbow(20)) +
  scale_y_reverse(limits = c(6500, 2500)) +
  labs(
    title = "AMAZING PENGUIN ANALYSIS!!!",
    subtitle = "This proves penguins are DEFINITELY getting bigger*",
    caption = "*not actually tested"
  ) +
  theme(
    plot.background = element_rect(fill = "yellow", color = "red", size = 5),
    panel.background = element_rect(fill = "lavender"),
    panel.grid.major = element_line(color = "red", size = 1.5),
    plot.title = element_text(size = 24, color = "purple", face = "bold.italic",
                              hjust = 0.5),
    axis.title.y = element_text(size = 22, color = "darkgreen", face = "bold"),
    axis.text.x = element_text(size = 14, color = "red", angle = 60, hjust = 1),
    legend.background = element_rect(fill = "pink", color = "green", size = 2)
  ) +
  annotate("text", x = 50, y = 4000, label = "WOW!", size = 20,
           color = "red", alpha = 0.3, angle = 30, fontface = "bold")

ugly_plot
```

### Misleaders used:

1. **Inverted Y-axis** -- reversed scale misleads interpretation
2. **Rainbow colormap** -- not perceptually uniform, bad for colorblind readers
3. **Inappropriate geom** -- line chart for non-sequential data
4. **Cluttered annotations** -- "WOW!" adds zero information
5. **Terrible color scheme** -- yellow background, red grid, purple title
6. **Excessive legend styling** -- pink with green border
7. **Y-axis label** -- huge green bold text, no units

## Misleader Examples

### 1. Truncated Y-Axis

```{r truncated-axis}
bad_data <- tibble(
  category = c("Product A", "Product B"),
  sales = c(98, 102)
)

p_bad1 <- ggplot(bad_data, aes(category, sales, fill = category)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(limits = c(95, 105)) +
  labs(title = "MISLEADING: Truncated Y-axis",
       subtitle = "Looks like 5x difference, actually ~4%") +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme_minimal()

p_good1 <- ggplot(bad_data, aes(category, sales, fill = category)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(limits = c(0, 110)) +
  labs(title = "CORRECT: Y-axis from 0",
       subtitle = "True proportion visible") +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme_minimal()

p_bad1 + p_good1
```

### 2. Pie Chart vs Bar Chart

```{r pie-vs-bar}
pie_data <- tibble(
  category = LETTERS[1:6],
  value = c(23, 21, 19, 17, 12, 8)
)

p_bad2 <- ggplot(pie_data, aes(x = "", y = value, fill = category)) +
  geom_col(width = 1) +
  coord_polar("y") +
  labs(title = "MISLEADING: Pie Chart",
       subtitle = "Hard to compare similar values") +
  theme_void()

p_good2 <- ggplot(pie_data, aes(x = reorder(category, value), y = value,
                                fill = category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "BETTER: Ordered Bar Chart",
       subtitle = "Easy to compare values", x = "", y = "Value") +
  theme_minimal()

p_bad2 + p_good2
```

### 3. Rainbow vs Perceptually Uniform Colormap

```{r rainbow-vs-viridis}
p_bad3 <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_gradientn(colors = rainbow(7)) +
  labs(title = "BAD: Rainbow colormap") +
  theme_minimal()

p_good3 <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "GOOD: Viridis (perceptually uniform)") +
  theme_minimal()

p_bad3 + p_good3
```

## Chart Makeover: Before and After

```{r makeover, fig.width=14}
p_before <- ggplot(penguins_clean, aes(body_mass_g, flipper_length_mm)) +
  geom_point(color = "red", size = 5, shape = 17) +
  labs(title = "penguin data") +
  theme(
    plot.background = element_rect(fill = "lightyellow"),
    panel.background = element_rect(fill = "lightblue"),
    panel.grid = element_line(color = "red")
  )

p_after <- ggplot(penguins_clean, aes(body_mass_g, flipper_length_mm,
                                      color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, size = 0.8) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Penguin Body Mass vs Flipper Length",
       subtitle = "Species-specific linear trends",
       x = "Body Mass (g)", y = "Flipper Length (mm)", color = "Species") +
  theme_minimal() +
  theme(legend.position = "bottom")

p_before + p_after + plot_annotation(title = "Chart Makeover: Before & After")
```

## Conclusions

- Visualization misleaders can dramatically distort data interpretation.
- The most common offenders: truncated axes, pie charts, rainbow colormaps.
- Good visualization requires: proper scales, clean themes, colorblind-friendly palettes, and appropriate chart types.

```{r session}
sessionInfo()
```
