---
title: "Lab 2: Automated Graphical Exploratory Data Analysis (autoGEDA)"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(patchwork)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## 1. Dataset: TidyTuesday Valentine's Day Spending

We use a TidyTuesday dataset on historical Valentine's Day spending in the US.

```{r data-load}
tuesdata <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-02-13/historical_spending.csv"
)

glimpse(tuesdata)
summary(tuesdata)
```

## 2. Data Overview

```{r overview}
summary(tuesdata)

# Data dimensions
cat("Rows:", nrow(tuesdata), "\nColumns:", ncol(tuesdata), "\n")
cat("Missing values:", sum(is.na(tuesdata)), "\n")
```

## 3. Manual autoGEDA Approach

### Histograms for all numeric variables

```{r histograms, fig.height=8}
numeric_vars <- tuesdata |> select(where(is.numeric)) |> names()

plots_hist <- map(numeric_vars, function(var) {
  ggplot(tuesdata, aes(x = .data[[var]])) +
    geom_histogram(fill = "steelblue", color = "white", bins = 30) +
    labs(title = var) +
    theme_minimal() +
    theme(plot.title = element_text(size = 10))
})

wrap_plots(plots_hist, ncol = 3)
```

### Boxplots for all numeric variables

```{r boxplots}
plots_box <- map(numeric_vars, function(var) {
  ggplot(tuesdata, aes(y = .data[[var]])) +
    geom_boxplot(fill = "steelblue", alpha = 0.5) +
    labs(title = var) +
    theme_minimal() +
    theme(plot.title = element_text(size = 10))
})

wrap_plots(plots_box, nrow = 1)
```

### Bar plots for categorical variables

```{r barplots}
cat_vars <- tuesdata |> select(where(~is.character(.) | is.factor(.))) |> names()

if (length(cat_vars) > 0) {
  plots_bar <- map(cat_vars, function(var) {
    ggplot(tuesdata, aes(x = fct_infreq(.data[[var]]))) +
      geom_bar(fill = "coral") +
      labs(title = var, x = "") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
  })
  wrap_plots(plots_bar, ncol = 2)
}
```

### Correlation matrix

```{r correlation}
if (length(numeric_vars) >= 2) {
  cor_mat <- tuesdata |> select(all_of(numeric_vars)) |> cor(use = "complete.obs")
  cor_df <- cor_mat |>
    as.data.frame() |> rownames_to_column("var1") |>
    pivot_longer(-var1, names_to = "var2", values_to = "r")

  ggplot(cor_df, aes(var1, var2, fill = r)) +
    geom_tile(color = "white") +
    geom_text(aes(label = round(r, 2)), size = 3) +
    scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral",
                         midpoint = 0) +
    labs(title = "Correlation Matrix") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    coord_fixed()
}
```

### Missing values

```{r missing}
miss_counts <- colSums(is.na(tuesdata))
miss_df <- tibble(variable = names(miss_counts), missing = miss_counts) |>
  filter(missing > 0)

if (nrow(miss_df) > 0) {
  ggplot(miss_df, aes(x = reorder(variable, missing), y = missing)) +
    geom_col(fill = "coral") +
    coord_flip() +
    labs(title = "Missing Values by Variable", x = "", y = "Count") +
    theme_minimal()
} else {
  cat("No missing values in the dataset.\n")
}
```

## 4. Table Summary

```{r table-summary}
tuesdata |> head(10) |> knitr::kable(caption = "First 10 rows of the dataset")
```

## Conclusions

- **dlookr** provides quick diagnostic overviews, normality checks, and outlier detection.
- **Manual approach** gives full control over visualization aesthetics and layout.
- Missing data visualization (naniar) is essential for understanding data quality.
- The correlation matrix reveals relationships between spending categories.

```{r session}
sessionInfo()
```
