---
title: "Lab 3: GEDA + Analytic Questions"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(patchwork)
library(palmerpenguins)

conflicts_prefer(dplyr::filter, dplyr::select)

data(penguins)
penguins_clean <- penguins |> drop_na()
```

## Dataset: Palmer Penguins

```{r overview}
glimpse(penguins_clean)
```

## GEDA: Distribution Overview

```{r distributions, fig.height=8}
p_bill_len <- ggplot(penguins_clean, aes(bill_length_mm)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  theme_minimal() + labs(title = "Bill Length")

p_bill_dep <- ggplot(penguins_clean, aes(bill_depth_mm)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  theme_minimal() + labs(title = "Bill Depth")

p_flip <- ggplot(penguins_clean, aes(flipper_length_mm)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  theme_minimal() + labs(title = "Flipper Length")

p_mass <- ggplot(penguins_clean, aes(body_mass_g)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  theme_minimal() + labs(title = "Body Mass")

(p_bill_len + p_bill_dep) / (p_flip + p_mass)
```

```{r categorical}
p_species <- ggplot(penguins_clean, aes(species, fill = species)) +
  geom_bar(show.legend = FALSE) + theme_minimal()
p_island <- ggplot(penguins_clean, aes(island, fill = island)) +
  geom_bar(show.legend = FALSE) + theme_minimal()
p_sex <- ggplot(penguins_clean, aes(sex, fill = sex)) +
  geom_bar(show.legend = FALSE) + theme_minimal()

p_species + p_island + p_sex
```

## Elementary Questions

### 1. IDENTIFY: Average body mass of Gentoo penguins?

```{r q1-identify}
gentoo_mass <- penguins_clean |>
  filter(species == "Gentoo") |>
  summarise(mean_mass = mean(body_mass_g))

ggplot(penguins_clean, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  geom_hline(yintercept = gentoo_mass$mean_mass, linetype = "dashed", color = "red") +
  annotate("text", x = 3.3, y = gentoo_mass$mean_mass + 100,
           label = paste0("Gentoo mean: ", round(gentoo_mass$mean_mass), "g"),
           color = "red", size = 3.5) +
  labs(title = "IDENTIFY: Average body mass of Gentoo penguins",
       subtitle = paste0("Answer: ", round(gentoo_mass$mean_mass), " grams"),
       x = "Species", y = "Body Mass (g)") +
  theme_minimal()
```

### 2. LOCATE: Where are Chinstrap penguins found?

```{r q2-locate}
ggplot(penguins_clean, aes(x = island, fill = species)) +
  geom_bar(position = "dodge") +
  labs(title = "LOCATE: Where are Chinstrap penguins found?",
       subtitle = "Answer: Only on Dream island",
       x = "Island", y = "Count") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

### 3. COMPARE: Do males have longer flippers?

```{r q3-compare}
ggplot(penguins_clean, aes(x = sex, y = flipper_length_mm, fill = sex)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  labs(title = "COMPARE: Do males have longer flippers than females?",
       subtitle = "Answer: Yes, males have consistently longer flippers",
       x = "Sex", y = "Flipper Length (mm)") +
  scale_fill_manual(values = c("coral", "steelblue")) +
  theme_minimal()
```

### 4. RANK: Which species is heaviest?

```{r q4-rank}
ggplot(penguins_clean, aes(x = reorder(species, body_mass_g, FUN = median),
                           y = body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  labs(title = "RANK: Which species is heaviest?",
       subtitle = "Answer: Gentoo > Chinstrap > Adelie",
       x = "Species (ordered by median mass)", y = "Body Mass (g)") +
  theme_minimal()
```

### 5. CONNECT: Bill length vs bill depth?

```{r q5-connect}
ggplot(penguins_clean, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "CONNECT: Are bill length and depth related?",
       subtitle = "Within species: positive. Overall: Simpson's paradox (negative).",
       x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

### 6. DISTINGUISH: What makes Gentoo different?

```{r q6-distinguish, fig.height=8}
p_d1 <- ggplot(penguins_clean, aes(bill_length_mm, flipper_length_mm, color = species)) +
  geom_point(alpha = 0.6) + scale_color_brewer(palette = "Set2") +
  labs(x = "Bill Length", y = "Flipper Length") + theme_minimal()

p_d2 <- ggplot(penguins_clean, aes(bill_depth_mm, body_mass_g, color = species)) +
  geom_point(alpha = 0.6) + scale_color_brewer(palette = "Set2") +
  labs(x = "Bill Depth", y = "Body Mass") + theme_minimal()

p_d3 <- ggplot(penguins_clean, aes(species, flipper_length_mm, fill = species)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  labs(title = "Flipper Length") + theme_minimal()

p_d4 <- ggplot(penguins_clean, aes(species, body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE, alpha = 0.7) +
  labs(title = "Body Mass") + theme_minimal()

(p_d1 + p_d2) / (p_d3 + p_d4) +
  plot_annotation(title = "DISTINGUISH: What makes Gentoo different?")
```

Gentoo penguins are clearly separated: larger flippers, higher body mass, and longer but shallower bills compared to Adelie and Chinstrap.

## Synoptic Questions

### 7. GROUP: Do species form distinct clusters?

```{r q7-group}
ggplot(penguins_clean, aes(x = bill_length_mm, y = flipper_length_mm,
                           color = species, shape = species)) +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  labs(title = "GROUP: Do species form distinct clusters?",
       subtitle = "Yes, three clear clusters with 95% confidence ellipses",
       x = "Bill Length (mm)", y = "Flipper Length (mm)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

### 8. CORRELATE: Dependencies between measurements?

```{r q8-correlate}
cor_data <- penguins_clean |>
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) |>
  cor()

cor_long <- cor_data |>
  as.data.frame() |> rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r")

ggplot(cor_long, aes(var1, var2, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(r, 2)), size = 4) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "coral", midpoint = 0) +
  labs(title = "CORRELATE: Dependencies between measurements") +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed()
```

### 9. TRENDS: Body mass vs bill length?

```{r q9-trends}
ggplot(penguins_clean, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.5) +
  geom_smooth(method = "loess", color = "black", se = TRUE) +
  labs(title = "TRENDS: Body mass vs bill length",
       subtitle = "Positive trend overall with species-specific patterns",
       x = "Bill Length (mm)", y = "Body Mass (g)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

### 10. CYCLES: Periodic patterns over years?

```{r q10-cycles}
penguins_by_year <- penguins |>
  drop_na() |>
  group_by(year, species) |>
  summarise(mean_mass = mean(body_mass_g), n = n(), .groups = "drop")

ggplot(penguins_by_year, aes(x = year, y = mean_mass, color = species)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(title = "CYCLES: Do measurements re-occur periodically?",
       subtitle = "Only 3 years of data - insufficient to detect cycles",
       x = "Year", y = "Mean Body Mass (g)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

### 11. OUTLIERS: Any special cases?

```{r q11-outliers}
ggplot(penguins_clean, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_point(
    data = penguins_clean |>
      filter(body_mass_g > 6000 | flipper_length_mm > 230 | flipper_length_mm < 175),
    color = "red", size = 4, shape = 1, stroke = 2
  ) +
  labs(title = "OUTLIERS: Any special observations?",
       subtitle = "Red circles mark extreme values",
       x = "Body Mass (g)", y = "Flipper Length (mm)") +
  scale_color_brewer(palette = "Set2") +
  theme_minimal()
```

### 12. FEATURES: What characterizes the data?

```{r q12-features}
ggplot(penguins_clean, aes(x = flipper_length_mm, fill = species)) +
  geom_density(alpha = 0.5) +
  labs(title = "FEATURES: What is characteristic of the data?",
       subtitle = "Bimodal distribution - Gentoo clearly separated from others",
       x = "Flipper Length (mm)", y = "Density") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Conclusions

- Palmer Penguins data shows **three distinct species clusters** in morphological space.
- **Simpson's paradox** appears in bill length vs depth (negative overall, positive within species).
- **Gentoo** penguins are clearly the largest species with the longest flippers.
- Only 3 years of data is insufficient for cycle detection.

```{r session}
sessionInfo()
```
