---
title: "Lab 6: Visualization Recreation (USAFacts)"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Task: Recreate USAFacts Foodborne Illness Chart

Group 4 was assigned the CDC foodborne illness pathogens visualization from USAFacts. The chart shows estimated annual foodborne illnesses from the top six pathogens with 90% credible intervals.

## Data

```{r data}
pathogens <- read_csv("../data-I2Rsw.csv") |>
  rename(
    pathogen = Pathogen,
    illnesses = Illnesses,
    low = `Low Estimate`,
    high = `High estimate`
  ) |>
  mutate(
    pathogen = fct_reorder(pathogen, illnesses),
    illnesses_label = case_when(
      illnesses >= 1e6 ~ paste0(round(illnesses / 1e6, 1), "M"),
      illnesses >= 1e3 ~ paste0(round(illnesses / 1e3, 1), "K"),
      TRUE ~ as.character(illnesses)
    ),
    low_label = case_when(
      low >= 1e6 ~ paste0(round(low / 1e6, 1), "M"),
      low >= 1e3 ~ paste0(round(low / 1e3, 1), "K"),
      TRUE ~ as.character(low)
    ),
    high_label = case_when(
      high >= 1e6 ~ paste0(round(high / 1e6, 1), "M"),
      high >= 1e3 ~ paste0(round(high / 1e3, 1), "K"),
      TRUE ~ as.character(high)
    )
  )

pathogens |> select(pathogen, low, illnesses, high)
```

## Recreation: Dumbbell Chart

```{r recreation, fig.width=10, fig.height=6}
usafacts_orange <- "#E87722"
usafacts_gray <- "#666666"
usafacts_lightgray <- "#D9D9D9"

ggplot(pathogens, aes(y = pathogen)) +
  geom_segment(aes(x = low, xend = high, yend = pathogen),
               color = usafacts_lightgray, size = 2) +
  geom_point(aes(x = low), color = usafacts_orange, size = 3) +
  geom_point(aes(x = illnesses), color = usafacts_orange, size = 5) +
  geom_point(aes(x = high), color = usafacts_orange, size = 3) +
  geom_text(aes(x = low, label = low_label), hjust = 1.3, size = 3,
            color = usafacts_gray) +
  geom_text(aes(x = illnesses, label = illnesses_label), vjust = -1.2,
            size = 3, color = usafacts_gray, fontface = "bold") +
  geom_text(aes(x = high, label = high_label), hjust = -0.3, size = 3,
            color = usafacts_gray) +
  scale_x_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "M"),
    breaks = seq(0, 10e6, 2e6),
    limits = c(-500000, 11000000)
  ) +
  labs(
    title = "Norovirus is the top pathogen contributing to foodborne\nillness, according to CDC estimates.",
    subtitle = "Estimated ranges of annual foodborne illnesses from the top six\npathogens, 2019 CDC study",
    x = "", y = "",
    caption = "Sorted by CDC midrange estimate. Ranges are 90% credible intervals.\nSource: CDC"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", lineheight = 1.2),
    plot.subtitle = element_text(size = 10, color = usafacts_gray,
                                 margin = margin(t = 5, b = 15)),
    plot.caption = element_text(size = 8, color = usafacts_gray, hjust = 0),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Alternative: Bar Chart with Error Bars

```{r bar-alternative}
ggplot(pathogens, aes(x = pathogen, y = illnesses, fill = pathogen)) +
  geom_col(show.legend = FALSE, fill = usafacts_orange, alpha = 0.8) +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0.3,
                color = usafacts_gray) +
  coord_flip() +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  labs(title = "Foodborne Illness Estimates with Uncertainty",
       subtitle = "Error bars show 90% credible intervals",
       x = "", y = "Estimated Annual Illnesses") +
  theme_minimal()
```

## Alternative: Log Scale Dot Plot

```{r log-alternative}
ggplot(pathogens, aes(x = illnesses, y = pathogen)) +
  geom_linerange(aes(xmin = low, xmax = high), color = "gray70", size = 1) +
  geom_point(size = 4, color = usafacts_orange) +
  scale_x_log10(labels = scales::label_number(scale = 1e-3, suffix = "K")) +
  labs(title = "Foodborne Illness Estimates (Log Scale)",
       subtitle = "Log scale reveals enormous range differences between pathogens",
       x = "Estimated Annual Illnesses (log scale)", y = "") +
  theme_minimal()
```

## camcorder Demo (Code Only)

The `camcorder` package records ggplot construction steps as a GIF:

```{r camcorder, eval=FALSE}
library(camcorder)

gg_record(dir = "recording", device = "png", width = 10, height = 6,
          units = "in", dpi = 300)

# Step 1: Basic points
ggplot(pathogens, aes(y = pathogen, x = illnesses)) +
  geom_point() + theme_minimal()

# Step 2: Add range segments
ggplot(pathogens, aes(y = pathogen)) +
  geom_segment(aes(x = low, xend = high, yend = pathogen),
               color = "gray80", size = 2) +
  geom_point(aes(x = illnesses), color = usafacts_orange, size = 5) +
  theme_minimal()

# Step 3: Full styled version (final plot above)

gg_playback(name = "build_animation.gif",
            first_image_duration = 2, last_image_duration = 5,
            frame_duration = 1, image_resize = 800)
```

## Conclusions

- The dumbbell/lollipop chart effectively shows point estimates with uncertainty ranges.
- Norovirus dominates with ~5.5M estimated annual cases.
- Alternative views (bar chart, log scale) offer different perspectives on the same data.

```{r session}
sessionInfo()
```
