---
title: "Lab 7: Tamara Munzner -- Foundations 1"
subtitle: "Visualization Methods in Data Analysis by R"
author: "Group 4, Variant 5"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(conflicted)
library(tidyverse)
library(palmerpenguins)

conflicts_prefer(dplyr::filter, dplyr::select)
```

## Part 1: Data Abstraction Framework

Munzner's data abstraction identifies four key questions:

1. **What** type of dataset? (table, network, spatial, set)
2. **What** are the attributes? (categorical, ordered, quantitative)
3. **Why** is the user looking at it? (discover, present, enjoy)
4. **How** should it be encoded? (marks and channels)

## Part 2: Palmer Penguins -- Data Abstraction

### Dataset type

**Flat table** -- 344 observations x 8 variables.

### Attribute types

| Variable | Type | Ordering |
|:---------|:-----|:---------|
| species | Categorical | Nominal |
| island | Categorical | Nominal |
| bill_length_mm | Quantitative | Sequential |
| bill_depth_mm | Quantitative | Sequential |
| flipper_length_mm | Quantitative | Sequential |
| body_mass_g | Quantitative | Sequential |
| sex | Categorical | Nominal |
| year | Quantitative | Sequential |

### Demonstration

```{r penguins-abstraction}
data(penguins)
penguins_clean <- penguins |> drop_na()

ggplot(penguins_clean, aes(x = bill_length_mm, y = flipper_length_mm,
                           color = species, size = body_mass_g)) +
  geom_point(alpha = 0.6) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Palmer Penguins: Multi-channel Encoding",
    subtitle = "Position (x,y), Color (species), Size (body mass)",
    x = "Bill Length (mm)", y = "Flipper Length (mm)",
    size = "Body Mass (g)"
  ) +
  theme_minimal()
```

**Marks**: Points (zero-dimensional).
**Channels**: Horizontal position (bill length), vertical position (flipper length), color hue (species), size (body mass).

## Part 3: Foreign Aid Dataset

Foreign aid data typically involves:

- **Dataset type**: Table (with potential network structure between donor--recipient)
- **Items**: Country-year pairs
- **Attributes**: Aid amount (quantitative), donor/recipient (categorical), year (ordered), sector (categorical)

### Key design decisions

1. **Task**: Compare aid amounts across donors and recipients over time
2. **Encoding**: Position on common scale (bar chart) is most effective for quantitative comparison
3. **Faceting**: By donor or region to manage complexity
4. **Color**: Sequential palette for amounts, qualitative for categories

```{r foreign-aid-example}
# Simulated foreign aid data for illustration
set.seed(42)
aid_data <- expand_grid(
  donor = c("USA", "UK", "Germany", "France", "Japan"),
  year = 2015:2020
) |>
  mutate(
    aid_billion = case_when(
      donor == "USA" ~ rnorm(n(), 35, 3),
      donor == "UK" ~ rnorm(n(), 18, 2),
      donor == "Germany" ~ rnorm(n(), 24, 2.5),
      donor == "France" ~ rnorm(n(), 12, 1.5),
      donor == "Japan" ~ rnorm(n(), 15, 2)
    )
  )

ggplot(aid_data, aes(x = year, y = aid_billion, color = donor)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set2") +
  labs(title = "Foreign Aid by Donor Country (Simulated)",
       x = "Year", y = "Aid (Billion USD)", color = "Donor") +
  theme_minimal()
```

## Munzner's What-Why-How Framework Summary

| Component | Question | Penguins Example |
|:----------|:---------|:-----------------|
| **What** | Data type? | Flat table, 8 attributes |
| **Why** | User task? | Discover clusters, compare species |
| **How** | Encoding? | Scatterplot with color + size channels |

## Conclusions

- Data abstraction is the first step in visualization design.
- Identifying attribute types (categorical vs. quantitative) determines which channels are appropriate.
- Munzner's framework provides a systematic approach to visualization design.

```{r session}
sessionInfo()
```
